---
title: "ClinicalMetadata"
output: html_notebook

---

# INIT
```{r}
source("./scripts/startup.R")
```
# MODELING STAGE

Demonstrate whether a linear regression can be fit or not
# INSERT LASSO DISTRIBUTION FOR LOG2 ON z transformed CONT. INPUTS

```{r}
#p_sub<- p %>% left_join(mcov_samples_filtered) #to add info about coverage

lasso_xy = function(x, y, family = "gaussian") {
  cv_model <- cv.glmnet(x, y, alpha = 1, family = family, nfolds=100)
  plot(cv_model) 
  best_model <- glmnet(x, y, alpha = 1, family = family, lambda = cv_model$lambda.min)
  best_model$dev.ratio
  lasso_coef = coef(best_model) %>% as.matrix() %>% data.frame() %>% rownames_to_column() %>% 
  select(factor=rowname, coefficient=s0) %>% filter(factor!="(Intercept)") %>% 
  arrange(desc(coefficient))
  return(lasso_coef)
}

normalize <- function(x) {

  scaled = (x- min(x)) /(max(x)-min(x))
    return(scaled)
}

```




# LASSO scan the features
Looks like it's important to remove the HIV and transplant because of sparsity of data.

```{r}
scan_factors = c('Duration','age18under','age55plus','sex','chronic_lung_disease',
                 'chronic_liver_disease', 'chronic_kidney_disease', 'chronic_heart_disease', 
                 'transplant_patient', 'hiv', 'hypertension', 'diabetes', 'cancer', 'obesity', 
                 'plasma', 'mAb', 'admitted_hospital','vaccine_status',
                 'vocAlpha','vocDelta','collection_month',
                 'surveillance','CT','median_coverage','run', 'PUI')

scale_scan_factors = function(patient_var_tmp, scan_factors) {
  p_sub = patient_var_tmp %>% 
  select(MCoVNumber,lineage,Duration,COLLECTION_DT:last_col(),one_of(scan_factors)) %>% 
  unique %>% as.data.frame()

  p_sub[, colnames(p_sub) %in% scan_factors] 
  p_sub_scaled = p_sub %>% select(one_of(scan_factors)) %>% 
    mutate_if(is.numeric, scale) %>% # scale will z scale it, instead of the normalize fx above which is min max
    cbind(p_sub %>% select(!one_of(scan_factors)))
  return(p_sub_scaled)
}

patient_var_tmp = read_feather("processing/patient_var_tmp.arrow")
scan_factors_trim = scan_factors[!scan_factors %in% c("transplant_patient", "hiv")] 
p_sub_scaled = scale_scan_factors(
  patient_var_tmp %>% filter(n_var < 30 & CT < 26), 
  scan_factors_trim)
y<-log2(p_sub_scaled$n_var+1)
x<-(p_sub_scaled[, colnames(p_sub_scaled) %in% scan_factors_trim]) %>% data.matrix()
lasso_under_30 = lasso_xy(x,y) %>% mutate(cutoff = "n_var_under_30") %>%
  arrange(coefficient) %>% mutate(factor = fct_reorder(factor, coefficient))

lasso_initial_features = lasso_under_30 %>% ggplot(aes(x=factor, y = coefficient)) + 
  geom_bar(stat="identity") + coord_flip()
lasso_initial_features
ggsave("ggsave/lasso_initial_features.pdf", lasso_initial_features, height = 4, width = 3)
```

```{r}
admitted_matrix = cbind.data.frame(x,#[x[,"admitted_hospital"] == 1,], 
                        n_var = normalize(p_sub_scaled$n_var))
clinical_cor_heatmap = pheatmap(abs(cor(admitted_matrix, method = "spearman")),
                                color = viridis(10), border_color	= NA, treeheight_row = 0)

ggsave("ggsave/clinical_cor_heatmap.pdf", clinical_cor_heatmap, height = 6, width = 5)
```

# Demonstrate which features have enough n to be scanned
```{r, fig.width = 8, fig.height = 4}
check_levels = function(x) { return(levels(x) %>% length ==2) }
numeric1 = function(x) { return(as.numeric(x)-1) }

plot_feature_n = function(p_sub_scaled) {
  sum_n = nrow(p_sub_scaled)
  feature_stats_IP = p_sub_scaled %>% select(one_of(scan_factors)) %>% 
    filter(admitted_hospital == 1) %>%
    select_if(is.factor) %>% select_if(check_levels) %>%
    mutate_if(is.factor, numeric1) %>% colSums() %>% 
    data.frame(feature = names(.), IP = .)
    sum_n_IP = feature_stats_IP["admitted_hospital","IP"]

  
  feature_stats_OP = p_sub_scaled %>% select(one_of(scan_factors)) %>% 
    filter(admitted_hospital == 0) %>%
    select_if(is.factor) %>% select_if(check_levels) %>%
    mutate_if(is.factor, numeric1) %>% colSums() %>% 
    data.frame(feature = names(.), OP = .)
  sum_n_OP = sum_n - sum_n_IP
  
  feature_stats = left_join(feature_stats_IP, feature_stats_OP)
  slope = sum_n_OP/sum_n_IP
  feature_n_plot = feature_stats %>% ggplot(aes(x = IP, y = OP, label = feature)) + 
    geom_abline(slope = slope, color = "red", linetype = "dashed") + 
    geom_point() + geom_text_repel() +
    scale_y_continuous( breaks = seq(0,1500,100), limits = c(0,1400),
                       sec.axis = sec_axis(~ . / sum_n_OP, labels = scales::label_percent(),
                                           breaks = seq(0,1,0.1))) + 
    scale_x_continuous( breaks = seq(0,2000,100), guide = guide_axis(angle = 90),
                       sec.axis = sec_axis(~ . / sum_n_IP, labels = scales::label_percent(),
                                           breaks = seq(0,1,0.1))) + theme_pubr() +
    xlab(paste0("Inpatient (n & % of ", sum_n_IP, ")")) +
    ylab(paste0("Outpatient (n & % of ", sum_n_OP, ")"))
  return(feature_n_plot)
}

undated_feature_n_plot = plot_feature_n(p_sub_scaled %>% filter(collection_date<="2021-07-01"))
julyonwards_feature_n_plot = plot_feature_n(p_sub_scaled %>% 
                                          filter(collection_date>="2021-07-01"))
feature_n_plot = ggarrange(undated_feature_n_plot + ggtitle("12/2020 - 07/2021"), 
          julyonwards_feature_n_plot + ggtitle("07/2021 - 11/2021"), align = "h",
          labels = list("A","B"))
feature_n_plot
ggsave("ggsave/feature_n_plot.pdf", feature_n_plot, width = 16, height = 8)
```

# PLOTTING COMMANDS PER CT

EXPLORATION OF THE ADMITTED

```{r}
patient_counts_30 = read_feather("processing/patient_counts_30.arrow")
patient_counts_30 = patient_counts_30 %>% #filter(collection_date >="2021-07-01") %>%
  mutate(admitted_hospital = as.factor(admitted_hospital))

ct_date_plot = function(patient_counts_30, admitted_hospital = "admitted_hospital") {
  plot_admitted = patient_counts_30 %>% 
    ggplot(aes(x = INSTRUMENT_RESULT, y = n_var, color = !!sym(admitted_hospital))) + 
    geom_point(shape = "") +
          geom_density_2d(aes(color = !!sym(admitted_hospital))) + 
    stat_cor(method = "spearman", cor.coef.name = "rho") +
      geom_smooth(se=TRUE) + theme_pubr() + scale_y_continuous(trans="log1p", 
        breaks = c(0,1,2,5,10,15,20,25,30)) +
    scale_x_continuous(limits = c(10,26))
  plot_admitted_marginal_by_ct = ggMarginal(plot_admitted, type = "violin", 
                                            draw_quantiles = c(0,0.25,0.5, 0.75),
             groupColour = TRUE, groupFill = TRUE)

# By Date now:
  plot_admitted_date = patient_counts_30 %>% group_by(admitted_hospital) %>% 
    ggplot(aes(x = collection_date, y = n_var, color = !!sym(admitted_hospital))) + 
    geom_point(shape = "") +
          geom_density_2d(aes(color = !!sym(admitted_hospital))) + 
    stat_cor(method = "spearman", cor.coef.name = "rho") +
      geom_smooth(se = TRUE) + theme_pubr() + scale_y_continuous(trans="log1p", 
                                                               breaks = c(0,1,2,5,10,15,20,25,30))
  plot_admitted_date_marginal = ggMarginal(plot_admitted_date,  groupColour = TRUE, groupFill = TRUE)
  
  plot_admitted_marginal_combined = ggarrange(plot_admitted_marginal_by_ct, 
                                              plot_admitted_date_marginal, align = "h")
}

plot_admitted_marginal_combined = ct_date_plot(patient_counts_30, "admitted_hospital")
plot_admitted_marginal_combined

ggsave("ggsave/plot_admitted_marginal_combined.pdf", 
       plot_admitted_marginal_combined, height = 4.5, width = 9)
```

OTHER FACTORS


```{r}
# VACCINATION GROUPS HISTOGRAM
vaccine_by_date_hist = patient_counts_30 %>% select(collection_month, vaccine_status, admitted_hospital) %>% 
  filter(vaccine_status == 1) %>% ggplot(aes(x = collection_month)) + 
  geom_histogram(stat="count") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        plot.background = element_rect(colour = "black", fill=NA, linewidth =2)) +
  ylab("vaccine count")
vaccine_by_date_hist

ggsave("ggsave/vaccine_by_date_hist.pdf", vaccine_by_date_hist, height = 2, width = 2)

# ADMITTED GROUPS HISTOGRAM
patient_counts_30 %>% select(collection_month, vaccine_status, admitted_hospital) %>%
  filter(admitted_hospital == 1) %>% ggplot(aes(x = collection_month)) + 
  geom_histogram(stat="count")

# Load in the TOTAL HISTOGRAM
lineages_histogram = readRDS("ggsave/lineages_figure.rds")
# combine the histograms
n_vax_admit = patient_counts_30 %>% select(COLLECTION_DT, vaccine_status, admitted_hospital) %>%
   mutate(month = floor_date(COLLECTION_DT, "month")) %>% group_by(month) %>%
  summarize(vaccinated = sum(as.numeric(as.character(vaccine_status))), 
            hospitalized = sum(as.numeric(as.character(admitted_hospital)))) %>%
   pivot_longer(!month, names_to = "total", values_to = "count")

lineages_histogram_vax_hosp = lineages_histogram +  
  geom_line(data = n_vax_admit, aes(month, count, color = total, linetype = total)) +
  scale_color_manual(values = c("grey10", "black")) + theme(legend.position = "right")

lineages_histogram_vax_hosp
# save the output
ggsave("ggsave/lineages_histogram_vax_hosp.pdf", lineages_histogram_vax_hosp, height = 4, width = 4)
```


```{r}
# CT VACCINATED
plot_vaccinated_marginal_by_ct = ct_date_plot(patient_counts_30 %>%
                                                filter(collection_date >="2021-07-01" & admitted_hospital == 1), "vaccine_status")
plot_vaccinated_marginal_by_ct

ggsave("ggsave/plot_vaccinated_marginal_by_ct.pdf", plot_vaccinated_marginal_by_ct, height = 5, width = 5)
```
```{r}
# cancer
plot_cancer_marginal_by_ct = ct_date_plot(patient_counts_30 %>%
                                                filter(collection_date <="2021-07-01"), "cancer")
plot_cancer_marginal_by_ct

test = ct_date_plot(patient_counts_30, "cancer")
test

```




```{r}
# VACCINATION GROUPS WITH CONTROL OVER THE TIME & OUTPATIENT STATUS
plot_vaccinated = patient_counts_30 %>%
  filter(collection_date>="2021-07-01") %>% 
  filter(admitted_hospital == 1) %>% 
  group_by(mAb) %>% 
  ggplot(aes(x = INSTRUMENT_RESULT, y = n_var, color = mAb)) + 
  geom_point(shape = "") +
        geom_density_2d(aes(color = mAb)) + stat_cor(method = "spearman", cor.coef.name = "rho") +
    geom_smooth(se=TRUE) + theme_pubr() + scale_y_continuous(trans="log1p", 
                                                             breaks = c(0,1,2,5,10,15,20,25,30)) +
    scale_x_continuous(limits = c(10,26))

    plot_vaccinated_marginal_by_ct = ggMarginal(plot_vaccinated, type = "boxplot", 
                                            draw_quantiles = c(0,0.25,0.5, 0.75),
             groupColour = TRUE, groupFill = TRUE)
    
plot_vaccinated_marginal_by_ct

#ggsave("ggsave/plot_mAb_marginal_by_ct.pdf", plot_vaccinated_marginal_by_ct, height = 5, width = 5)
```


# Healthcare workers
```{r}
# HCW GROUPS WITH CONTROL OVER THE TIME & OUTPATIENT STATUS
table(patient_counts_30 %>%  select(admitted_hospital,surveillance))

plot_hcw= patient_counts_30 %>% filter(admitted_hospital == 0) %>% filter(collection_date>="2021-07-01") %>%
  group_by(surveillance) %>% 
  ggplot(aes(x = INSTRUMENT_RESULT, y = n_var, color = vaccine_status)) + 
  geom_point(shape = "") +
        geom_density_2d(aes(color = vaccine_status)) + 
  stat_cor(method = "spearman", cor.coef.name = "rho") +
    geom_smooth(se=TRUE) + theme_pubr() + scale_y_continuous(trans="log1p", 
                                                             breaks = c(0,1,2,5,10,15,20,25,30))
((
  plot_hcw_marginal_by_ct = ggMarginal(plot_hcw, type = "violin", 
                                            draw_quantiles = c(0,0.25,0.5, 0.75),
             groupColour = TRUE, groupFill = TRUE)
))
```